Stochastic Approximation Monte Carlo with a Dynamic Update Factor (SAD)

### 2019-05-09 papers/histogram/sad.tex
start editing sad paper

\section{Introduction}
Over the past several decades a number of flat histogram Monte-Carlo
simulation algorithms have been developed which calculate the
thermodynamic properties of various systems for all temperatures.  The
development began with the original histogram method, which used a
single canonical Monte Carlo simulation to predict properties for
nearby temperatures~\cite{ferrenberg1988new}.  For large systems this
approach is limited to a narrow temperature range because a single
canonical simulation explores only a small range of energies.  This
led to a variety of ``broad'' (or ``flat'') histogram
methods~\cite{penna1996broad, penna1998broad, swendsen1999transition,
  wang2001determining, wang2001efficient, trebst2004optimizing}, which
attempt to explore a wide range of energies.  Another benefit, in
contrast with low-temperature canonical Monte Carlo, is that these
approaches cannot be trapped in a local energy minimum.

Wang and Landau introduced an algorithm (WL) that uses an update factor
and a statistical histogram to compute the density of states of a given
system~\cite{wang2001determining, wang2001efficient}.  While the method
is incredibly powerful, it has a few disadvantages. Firstly, it
requires the user to select a number of parameters, including the range
of energies to be studied~\cite{landau2004new}. This adds an additional
hurdle to its application to a variety of systems because the user must
simulate until the energy range of interest is known.  The simulation
violates detailed balance albiet briefly as the size of the violation decreases
with time, which complicates convergence analysis.  In fact, the error
in a WL computation the error has been demonstrated to saturate at a
non-zero value~\cite{yan2003fast}, i.e. the method does not converge to
the true density of states~\cite{belardinelli2007wang,
belardinelli2007fast, belardinelli2008analysis,
belardinelli2014intrinsic, singh2012density}.
%
Belardinelli and Pereyra demonstrated that allowing the update factor
to decrease faster than $1/t$ leads to
nonconvergence~\cite{belardinelli2007wang}.
%% Around the same time, the convergence rate for WL was being
%% explored~\cite{zhou2005understanding,lee2006convergence,
%%   belardinelli2007wang}.
%
This leads to their $1/t$-WL algorithm which ensures that the error
continues to decrease asymptotically as
$1/\sqrt{t}$~\cite{belardinelli2008analysis}.

Liang, Liu, and Carrol began to consider that WL could be considered
a special case of Stochastic Approximation whose convergence could be
mathematically proven~\cite{liang2006theory, liang2007stochastic}. In
2007, Liang et al.~\cite{liang2007stochastic} argued that WL can be
considered a form of Stochastic Approximation Monte Carlo (SAMC). While
SAMC can guarantee convergence, the method still has a system specific
user-defined variable which makes applying this algorithm to arbitrary
systems difficult.

>>>>

\section{Introduction}
Over the past several decades a number of flat histogram Monte-Carlo
simulation algorithms have been developed which calculate the
thermodynamic properties of various systems for over a range of temperatures.  This
development began with the original histogram method, which used a
single canonical Monte Carlo simulation to predict properties for
nearby temperatures~\cite{ferrenberg1988new}.  For large systems this
approach is limited to a narrow temperature range because a single
canonical simulation explores only a small range of energies.  This
led to a variety of ``flat'' (or ``broad'') histogram
methods~\cite{penna1996broad, penna1998broad, swendsen1999transition,
  wang2001determining, wang2001efficient, trebst2004optimizing}, which
explore a wider range of energies.  Another benefit, in
contrast with low-temperature canonical Monte Carlo, is that these
approaches cannot be trapped in a local energy minimum.

Wang and Landau introduced the most widely used flat histogram
algorithm (WL) that uses an update factor
and a statistical histogram to compute the density of states of a given
system~\cite{wang2001determining, wang2001efficient}.  While the method
is incredibly powerful, it has a few disadvantages. Firstly, it
requires the user to select the range
of energies to be studied~\cite{landau2004new}. This adds an additional
hurdle to its application to systems for which the interesting range
of energies is not known \emph{a priori}.  The simulation
violates detailed balance albiet briefly as the size of the violation decreases
with time, which complicates convergence analysis.  In fact, the error
in a WL computation the error has been demonstrated to saturate at a
non-zero value~\cite{yan2003fast}, i.e. the method does not converge to
the true density of states~\cite{belardinelli2007wang,
belardinelli2007fast, belardinelli2008analysis,
belardinelli2014intrinsic, singh2012density}.
%
Belardinelli and Pereyra demonstrated that allowing the update factor
to decrease faster than $1/t$ leads to
nonconvergence~\cite{belardinelli2007wang}.
%% Around the same time, the convergence rate for WL was being
%% explored~\cite{zhou2005understanding,lee2006convergence,
%%   belardinelli2007wang}.
%
This leads to their $1/t$-WL algorithm which ensures that the error
continues to decrease asymptotically as
$1/\sqrt{t}$~\cite{belardinelli2008analysis}.

Liang began to consider that WL could be considered
a special case of the Stochastic Approximation approach, whose
convergence could be
mathematically proven~\cite{liang2006theory}. In
2007, Liang, Liu, and Carrol et al. argued that WL can be
considered a form of Stochastic Approximation Monte Carlo
(SAMC)~\cite{liang2007stochastic}. \davidsays{I'm here} While
SAMC can guarantee convergence, the method still has a system specific
user-defined variable which makes applying this algorithm to arbitrary
systems difficult.
